Wait-free Queues with Polylogarithmic Step Complexity

### 2022-10-11 thesis/thesis.tex
update refrences

\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would no be poly-logarithmic time.
Instead we use implicit representation of the operations propagated together. Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}). Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}, a piece of information about the operations in a \nf{Refresh} step. A block contain the number of operations from the left and the right child in a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example). A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Furthermore, since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

\paragraph{Introduce Get and Index} So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ we can apply all the operations till $op$ in the root on a sequential instance of the object and then return the response. But Having this agreement by itself is not enough for an efficient queue.
A process may wish to know (1) the $i$th propagated operation or (2) the rank of a propagated operation in the linearization. We will explain how to use (1) and (2) to construct a queue in the next paragraph.

\paragraph{Explain pieces of a block helping to do Get and Index}
After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). Each block stores an extra constant amount of information (like prefix sums) to allow binary search to find the required block in a node quickly

To get the $i$th operation in the root, we find the block $B$ containing $i$th operation in the root, and then recursively find the subblock of $B$ that contains $i$th element. To make this recursive search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search. We also store pointers to determine the range of subblocks of a block to make binary search faster. In each block, we store the prefix sum of operations from the left child and the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in. After a block is installed we set the approximate index of the superblock of the block in it.

>>>>

\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be poly-logarithmic time.
Instead we use implicit representation of the operations propagated together. Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}). Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block is a piece of information about the operations in a \nf{Refresh} step. It contains the number of operations from the left and the right child in a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example). A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parents' blocks.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently which means there are less than or equal $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

\paragraph{Introduce Get and Index} So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ by applying all the operations till $op$ in the root on a sequential instance of the object and then return the response. Having this agreement by itself is not enough for an efficient queue.
A process may wish to know (1) the $i$th propagated operation or (2) the rank of a propagated operation in the linearization. We will explain how to use (1) and (2) to construct a fast queue in the next paragraph.

\paragraph{Explain pieces of a block helping to do Get and Index}
After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing $i$th operation in the root, and then recursively find the subblock of $B$ that contains $i$th element. To make this recursive search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search. We also store pointers to determine the range of subblocks of a block to make binary search faster. In each block, we store the prefix sum of operations from the left child and the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in. After a block is installed we set the approximate index of the superblock of the block in it.
